---
title: "STAT 578 - Advanced Bayesian Modeling - Fall 2019  Assignment 4"
author: "Xiaoming Ji"
output:
  pdf_document: default
---
## Solution for Problem (a)
### (i)
According to the formular for Moore's law and take log on T, we get:

$log(T) \approx log(C * 2 ^{A/2}) \rightarrow log(T) \approx log(C) + 0.3465736 * A$

This is the form of linear regression that make logT as response variable and A as explanatory variable. `log(C)` is the intercept and coefficient of A is 0.3465736. Thus we can say logT is roughly follow a simple linear regression on A.

### (ii)
```{r}
moores_df = read.csv("mooreslawdata.csv", header=TRUE)
with(moores_df, plot(log(Transistors) ~ Year))
```

## Solution for Problem (b)

```{r}
moores_model = lm(log(Transistors) ~ Year, data = moores_df)
summary(moores_model)
```

We built a classical linear regression model using `Year` (for A) as explanatory variable and `log(Transistors)` (for logT) as response variable. The summary shows that this model acheived good R-squared value (> 0.95), thus we can say logT is roughly follow a simple linear regression on A. The coefficient of A is `r coef(moores_model)[['Year']]`

## Solution for Problem (c)
## Solution for Problem (d)
